FFT based convolution

### 2010-03-09 article/Article.tex
article continued
Ignore-this: 10cd3a5e71450026f96dd7c30dc2cf63

darcs-hash:20100309014443-2fc9d-edfd21b5f498036bd9e9a7a142bb1f7471c20687.gz

\subsection{Zero padding}

The images will be padded with zeros for several reasons:
\begin{itemize}
  \item to make their size match. This is required to perform the multiplication in the frequency
  domain.
  \item  to avoid the border effects. Because the FFT considere the image as a cyclic signal, the
  image must be padded to avoid the border effects. The final image size for all the dimensions
  must be at least $P+Q-1$, where $P$ is the size of the first image on a dimension and $Q$ the size
  of the other image on the same dimension.
  \item to make the FFT possible, or to enhance its performance. Some FFT implementations, like the
  VNL one, can only be run on an image where the size on all the dimensions are a power of two.
  Some other implementations, like FFTW's one, are performing very differently depending on the
  size of the image. Adding a few more pixels on the border of the image can lead them to perform
  a lot better, as shown in chapter \ref{perf:gpf}.
\end{itemize}

@@

The convolution is a flipping transform by nature: the shape of the kernel image can be found in the
convolved image, but flipped on all the axes. To get the same behavior when computing the product
in the frequency domain, the kernel image must be flipped prior to the FFT.

@@

The convolution is usually done with a centered kernel. If the kernel is not centered, the input image
is shifted in the convolved image, by the same shift than the shift of the kernel. However, for the
FFT, the center pixel is the one in the corner, not the one in the center of the image, as is it 
usual for a human. To avoid a shift in the convolved image, the pixels must be shifted to the corner
of the image as expeced by the FFT.

@@

The convolved image in the frequency domain is transformed to the space domain with an inverse Fourier
transform.

@@

The image is cropped to fit the size of the input image.

@@

\subsection{FFTZeroPaddingImageFilter}

The padding step, as described in the previous chapter, has several goals, which lead to a single
image size used to pad both the input image and the kernel.
All the logic is implemented in itk::FFTZeroPaddingImageFilter.

@@

The image to convolve is simply padded with zeros. The kernel, however, requires a little more work to be
centered in the padded image, and to make its region match the one of the padded image to convolve.
The padding and the region changes are implemented as a minipipeline of itk::ConstantPadImageFilter and
itk::ChangeInformationImageFilter.

itk::FFTZeroPaddingImageFilter is also able to extend the padded region in order to enhance the FFT
performance, or to make it possible when the FFT possible when it require a size which is a power of
two. FFTW is able to work on any image size, but due to the algorithms useds, produce significantly
more performant result with some specific sizes. The two main reasons are:
\begin{itemize}
  \item the decomposition of images of composite sizes in smaller transforms using the
  Cooley-Tukey algorithm;
  \item the hard coded loop unrolling for the size which are prime numbers up to thirteen.
\end{itemize}
For these reasons, FFTW performs better when the size on each dimension has its greatest prime
factor smaller or equal to thirteen.
The greatest prime factor can be simply computed with the algorithm described in algorithm \ref{greatestPrimeFactor}.

@@

@@

@@

\verb$itk::FFTConvolutionImageFilter$ is a convenient minipipeline filter which groups all the required
filters to make the FFT based convolution in a single and easy to use filter.
It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagate the
user specified number of threads.

It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTZeroPaddingImageFilter$ to let
the user choose between the extra performance added by the padding and the extra memory cost.

@@

\subsection{Greatest prime factor}
\label{perf:gpf}

@@

% \url{http://www.itk.org}
% \code{Insight/Documentation/Style.pdf}
% \section{Principles of Solar Spot Detection}
% \cite{ITKSoftwareGuide}.
% \doxygen{ImageToImageFilter}

% \small \begin{verbatim}
% \end{verbatim} \normalsize

% The {itemize} environment uses a bullet for each \item.  If you want the 
% \item's numbered, use the {enumerate} environment instead.
% \begin{itemize}
%   \item  Insight Toolkit 2.4.
%   \item  CMake 2.2
% \end{itemize}

% \ref{cthead1}

\begin{figure}[htbp]
\centering
\includegraphics{cthead1}
\caption{The input image.\label{cthead1}}
\end{figure}

>>>>

\subsection{Padding}

The image to convolve and the kernel will be padded with for several reasons:
\begin{itemize}
  \item to make their size match. This is required to perform the multiplication in the frequency
  domain.
  \item  to avoid the border effects. Because the FFT considere the image as a cyclic signal, the
  image must be padded to avoid the border effects. The final image size for all the dimensions
  must be at least $P+Q-1$, where $P$ is the size of the first image on a dimension and $Q$ the size
  of the other image on the same dimension.
  \item to make the FFT possible, or to enhance its performance. Some FFT implementations, like the
  VNL one, can only be run on an image where the size on all the dimensions are a power of two.
  Some other implementations, like FFTW's one, are performing very differently depending on the
  size of the image. Adding a few more pixels on the border of the image can lead them to perform
  a lot better, as shown in chapter \ref{perf:gpf}.
\end{itemize}

The kernel is always padded with zeros, but the image to convolve can be padded in several ways:
\begin{itemize}
 \item zeros -- while simple, this method may show darkened border in the convolved image;
 \item zero flux Neumann -- this method extends the pixels on the image border in the padded zone, and
       is quite efficient in keeping the border effect low;
 \item mirror -- the image is padded with a mirror of the image. This is also a quite good way to
       keep the border effect low;
 \item wrap -- the opposite border of the image is copied in the padded zone. This may be useful to
       keep the circular behavior of the FFT, while still padding to fit the size requirement of the
       FFT implementation.
\end{itemize}
The result of those padding can be seen in figure \ref{fig:padding}.

\begin{figure}[htbp]
\begin{center}
\subfigure[Zero]{\includegraphics[scale=0.4]{pad-image-2}}
\subfigure[Zero flux Neumann]{\includegraphics[scale=0.4]{pad-image-zero-flux}}
\subfigure[Mirror]{\includegraphics[scale=0.4]{pad-image-mirror}}
\subfigure[Wrap]{\includegraphics[scale=0.4]{pad-image-wrap}}
\subfigure[Input image]{\includegraphics[scale=0.4]{cthead1}}
\caption{Padding a 256 $\times$ 256 image to a 512 $\times$ 512 one in several ways.\label{fig:padding}}
\end{center}
\end{figure}

@@

The convolution is a flipping transform by nature: the shape of the kernel image can be found in the
convolved image, but flipped on all the axes. To get the same behavior when computing the product
in the frequency domain, the kernel image must be flipped prior to the FFT. See figure \ref{fig:kernel-modifications}.

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image (scaled 20 times)]{\includegraphics[scale=8]{kernel}}
\subfigure[Zero padded]{\includegraphics[scale=0.4]{pad-kernel-2}}
\subfigure[Zero padded and flipped]{\includegraphics[scale=0.4]{kernel-pad-flip}}
\subfigure[Zero padded, flipped and shifted]{\includegraphics[scale=0.4]{kernel-pad-flip-shift}}
\caption{Kernel modifications in order to perform the FFT based convolution. Normalization is
only a change in intensity, and thus is not shown here.\label{fig:kernel-modifications}}
\end{center}
\end{figure}

@@

The convolution is usually done with a centered kernel. If the kernel is not centered, the input image
is shifted in the convolved image, by the same shift than the shift of the kernel. However, for the
FFT, the center pixel is the one in the corner, not the one in the center of the image, as is it 
usual for a human. To avoid a shift in the convolved image, the pixels must be shifted to the corner
of the image as expeced by the FFT. See figure \ref{fig:kernel-modifications}.

@@

The convolved image in the frequency domain is transformed to the space domain with an inverse Fourier
transform. See figure \ref{fig:image-modifications}.

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=0.4]{cthead1}}
\subfigure[Zero flux Neumann padded]{\includegraphics[scale=0.4]{pad-image-zero-flux}}
\subfigure[Convolved, still padded]{\includegraphics[scale=0.4]{image-pad-convolved}}
\subfigure[Cropped to the input image size]{\includegraphics[scale=0.4]{fftconv}}
\caption{The modifications applied to the input image.\label{fig:image-modifications}}
\end{center}
\end{figure}

@@

The image is cropped to fit the size of the input image. See figure \ref{fig:image-modifications}.

@@

\subsection{ZeroFluxNeumannPadImageFilter}

The zero flux Neumann padding filter was not implemented in ITK. It has been implemented by modifying
the \verb$ConstantPadImageFilter$ code.

\subsection{FFTPadImageFilter}

The padding step, as described in the previous chapter, has several goals, which lead to a single
image size used to pad both the input image and the kernel.
All the logic is implemented in \verb$itk::FFTPadImageFilter$.

@@

The image to convolve is simply padded with one of the available method: zeros, zero flux Neumann, mirror
or wrap.

The kernel, however, requires a little more work to be
centered in the padded image, and to make its region match the one of the padded image to convolve.
The padding and the region changes are implemented as a minipipeline of \verb$itk::ConstantPadImageFilter$ and
\verb$itk::ChangeInformationImageFilter$.

\verb$itk::FFTPadImageFilter$ is also able to extend the padded region in order to enhance the FFT
performance, or to make it possible when the FFT possible when it require a size which is a power of
two. FFTW is able to work on any image size, but due to the algorithms useds, produce significantly
more performant result with some specific sizes. The two main reasons are:
\begin{itemize}
  \item the decomposition of images of composite sizes in smaller transforms using the
  Cooley-Tukey algorithm;
  \item the hard coded loop unrolling for the size which are prime numbers up to thirteen.
\end{itemize}
For these reasons, FFTW performs better when the size on each dimension has its greatest prime
factor smaller or equal to thirteen.
The greatest prime factor can be simply computed with the algorithm described in algorithm \ref{greatestPrimeFactor}.

@@

\verb$FFTPadImageFilter$ can also be used with a single input. This is useful to be able to run a FFT
on an image with a size which is not a power of 2 with the vnl implementation, or to enhance the performance
of the FFTW implementation, without necessairly doing a convolution.

@@

\subsubsection{Still to be done}

There are still some missing features in the FFTW filters. The ones which comes to mind are:
\begin{itemize}
 \item let the user create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
\verb$FFTW_EXHAUSTIVE$, so that FFTW can create better plan with \verb$FFTW_ESTIMATE$ in a
later run;
 \item provide an API to save the FFTW wisdoms.
\end{itemize}

@@

\verb$itk::FFTConvolutionImageFilter$ is a convenient minipipeline filter which groups all the required
filters to make the FFT based convolution in a single and easy to use filter. See figure \ref{fig:conv-pipeline}
for an overview of the minipipeline.
\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{conv}
\caption{Internal pipeline of FFTConvolutionImageFilter.\label{fig:conv-pipeline}}
\end{figure}

It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagate the
user specified number of threads.

It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTPadImageFilter$ to let
the user choose between the extra performance added by the padding and the extra memory cost, and
\verb$PadMethod$ from the same filter to let the user choose the padding method -- zero flux Neumann
is used by default.

Finally, the \verb$Normalize$ attribute let the user choose if the kernel must be normalized to one
or not.

@@

\subsection{Greatest prime factor}
\label{perf:gpf}
As visible in figure \ref{fig:perf-fftw3}, the FFTW library performs quite differently depending on the
size of the input. A good way to get the best performance seems to be to use the sizes with the greatest
prime factor smaller or equal to 13.

Most of the time, and despite the usual advice to use a size which is a power of two,
those sizes perform significantly better than the closest sizes which are a power of two, probably
because of the smaller impact of the number of pixels in the image.

\begin{figure}[htbp]
\begin{center}
\subfigure[Execution time for an image of size $dim^{3}$.]{\includegraphics[scale=0.7]{fftw3-time-dim}}
\subfigure[Effect of the image greatest prime factor of the image size (gpf) on the number of pixels
 processed per seconds (ops).]{\includegraphics[scale=0.7]{fftw3-ops-gpf}}
\caption{Effect of the image size on the FFTW performance. Red dots are the sizes with a greatest prime
factor smaller or equal to 13. The grey dots are the ones with a greatest prime factor greater than 13.
\label{fig:perf-fftw3}}
\end{center}
\end{figure}

@@

\section{Conclusion}

This contribution provides with an efficient implementation of the convolution transform for large kernels.
The FFTW filters implementations has been significantly enhanced, and a set of reusable tools is provided.

Some other work may be done on this basis -- an non exhaustive list may be:
\begin{itemize}
 \item a binary erosion/dilation based on FFT;
 \item a set of deconvolution filters;
 \item a more memory efficient implementation of the convolution filter using the Overlap-Save algorithm.
\end{itemize}
